---
title: "TidyTuesday: Astronomy Picture of the Day"
author: "Ian R. Sturgill"
format:
  html:
    title-block-banner: true
    theme: sandstone
toc: true
toc-location: left
editor: visual
execute:
  df-print: kable
  warning: false
  message: false
number-sections: true
---

# Introduction {#sec-intro}

From the TidyTuesday README for 2026-01-20:

This week we're exploring the Astronomy Picture of the Day (APOD) archive. APOD is a popular NASA website featuring daily astronomy related images with a scientific explanation.\
Each day a different image or photograph of our universe is featured, along with a brief explanation. This APOD archive contains image information from the 2007 - 2025, pulled together into the `astropic` R package.

- What types of objects are most common in the archive?
- Are any images posted more than once?

Thank you to [Erin Grand](https://github.com/eringrand) for curating this week's dataset.

# Data cleaning {#sec-cleaning}

## Load libraries {#sec-libraries}

Here, the original prepared data came from a package installed via GitHub. Alternatively, you can install with `tidytuesdayR`.

```{r}
# remotes::install_github("eringrand/astropic")
# Optional: install.packages("tidytuesdayR")
library(tidyverse)
library(ggsci)
library(astropic)
# library(tidytuesdayR)
```

## Load in original data {#sec-data}

Here we'll choose to work via the original `astropic` library to load the `hist_apod` astronomy data:

```{r}
# Load in data as a tibble
data("hist_apod")

hist_apod <- hist_apod |>
  as_tibble()

# Preview column data, truncating the very long explanations
hist_apod |> 
  mutate(explanation = str_trunc(explanation, width = 20)) |>
  head()
```

The author recommends removing the `service_version` column, which just contains the value `v1` for every row. Let's also filter out rows that don't have data we need to answer the two questions presented in the @sec-intro, specifically: `media_type` and `title`.

```{r}
# Remove unneeded service_version column
# Then, filter to remove rows with missing data needed for our analyses
hist_apod <- hist_apod |>
  select(-service_version) |>
  filter(!is.na(media_type) & !is.na(title))
```

## Assumptions {#sec-assumptions}

Now, we should be good to go, noting a couple of assumptions:

1.  The media type is accurately recorded and "other" is genuinely a non-image, non-video type, rather than a collection or some other overlap with those categories.
2.  If an image was posted more than once, it would have the exact same title each time it was posted.

# Data analyses {#sec-analyses}

Let's set a few variables for later plot labels:

```{r}
# URL for the 2026-01-20 dataset
github_link <- "https://github.com/rfordatascience/tidytuesday/blob/main/data/2026/2026-01-20"

tidy_tuesday <- str_c("TidyTuesday: ", github_link)
```

## Most common object types {#sec-object-types}

Let's first take a look at the most common object types (`media_type`):

```{r}
data_title <- "Most frequent media types in the Astronomy Picture of the Day Dataset"

# Show the tibble first
hist_apod |>
  summarize(num_media = n(), .by = media_type) |>
  arrange(desc(num_media)) |>
  head()

# Now plot the data
hist_apod |>
  summarize(num_media = n(), .by = media_type) |>
  arrange(desc(num_media)) |>
  ggplot(aes(x = reorder(media_type, -num_media), 
             y = num_media,
             fill = media_type)) +
  geom_col() +
  scale_fill_npg() +
  guides(fill = "none") +
  labs(title = data_title,
       subtitle = tidy_tuesday,
       x = NULL,
       y = "Number of media objects in dataset") +
  theme_bw()
```

Here we can see that the most common media type, vastly outnumbering the other media types, is "image," which makes sense – this is the Astronomy Picture of the Day dataset, after all.

## Most re-posted images {#sec-re-posted}

Now let's see what the most re-posted images are, keeping in mind assumption #2 (@sec-assumptions).

```{r}
data_title2 <- "Top 4 most-frequently posted images in the Astronomy Picture of the Day Dataset"

# Show the top 10 in the tibble
hist_apod |>
  summarize(num_title = n(), .by = title) |>
  arrange(desc(num_title)) |>
  head()

# Plot the top 4 (many ties at 8 re-postings)
# Wrap the title text so it doesn't overlap in the plot
hist_apod |>
  summarize(num_title = n(), .by = title) |>
  arrange(desc(num_title)) |>
  slice(1:4) |>
  ggplot(aes(x = reorder(str_wrap(title, width = 20), -num_title), 
             y = num_title,
             fill = title)) +
  geom_col() +
  scale_y_continuous(breaks = seq(0, 10, by = 1)) +
  scale_fill_npg() + 
  guides(fill = "none") +
  labs(title = data_title,
       subtitle = tidy_tuesday,
       x = NULL,
       y = "Number of times posted") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  theme_bw()
```

The winner of most-popular (most re-posted) image is "The Horsehead Nebula," at 11 posts.

# sessionInfo() {#sec-sessioninfo}

```{r}
sessionInfo()
```
